---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# survtabler

<!-- badges: start -->
<!-- badges: end -->

`survival` package in R is great for interactively and flexibly running survival models and evaluate the resulting model. However, it does not "teach" any structure for an efficient workflow, and often leads to imperative programming and long and complicated analysis scripts. The aim of `survtabler` is to make survival analysis fast, concise, and intuitive while keeping the analysis code easy to read. This is achieved by *function-based* workflow, where many typical steps (modeling, graphing, analyses of violations of proportional hazards model) are automated with functions.

## Installation

You can install the development version of `survtabler` from [GitHub](https://github.com/) with:

```{r example_1}
# install.packages("devtools")
# devtools::install_github("jkoskenniemi/survtabler")
```

## Example

The first and the most important step is to provide a recipe for the survival analysis in R. This is done by creating a `survtable` object: essentially a data.frame that lists in each row two items needed for Cox Proportional Hazards Model modeled under the hood using `survival::coxph()`: `formula` and `data`. Function `create_survtable()` lists all combinations of analysis data (`data_name`) and variables given for outcome, exposure, survival time (`outcome_vars`, `exposure_vars`, `time_var`). Using these input and `covariates` used to adjust the model, it formulates the Cox Porportional Hazards model `formula` for each combination (each row of the `survtable` data frame).

```{r example_2}
library(survtabler)
library(magrittr)

#Specify all combinations of exposure, outcome, time variables and data_name
survtable_1 <- create_survtable(exposure_vars = c("exposure_2cat", "exposure_continuous"),
                 outcome_vars = c("outcome1", "outcome2"),
                 covariates = "age + sex + hla",
                 time_var = "cens_time",
                 data_name = "example_ti")
```


Once the `survtable` is established, it is very easy to run Cox Proportional Hazards model for each combination and return the result as a list

```{r example_3}
models <- survtable_1 %>%  
  model_survtable()
```


And then get the model coefficients and draw a forest plot, ...
```{r example_4}
models  %>%  
  get_coefs(c("exposure_2cat", "exposure_continuous"))  %>%  #Get coefficients for forrest plots
  graph_coefs(title = "**Your title here**") #Draw forest plots
```

... other model metadata, ...
```{r example_5}
models  %>%  
  get_model_meta()
```

..., and finally analyze the models for departures of proportionality of the hazards assumption.
```{r example_6}
models  %>%  
  catch_nonph()
```

